Inside Out: Externalizing Assumptions in Data Analysis as Validation Checks

### 2024-12-24 _manuscript/_tex/index.tex
section 4

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While some researchers
may struggle to identify the root causes, experienced researchers can
often quickly diagnose problems by checking a few key assumptions. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
introduce the term \emph{analysis validation checks} to formalize and
externalize these informal assumptions. We then introduce a procedure to
identify a subset of checks that best predict the occurrence of
unexpected outcomes, based on simulations of the original data. The
checks are evaluated in terms of accuracy, determined by binary
classification metrics, and redundancy, which measures the shared
information among checks using mutual information. We demonstrate this
approach with a toy example using step count data and a generalized
linear model example examining the effect of particulate matter air
pollution on daily mortality.
\end{abstract}

@@

In this paper, we conceptualize these inexplicit expectations as
\emph{analysis validation checks}, which allows us to examine the
assumptions made about the data during an analysis. We then introduce a
procedure to identify the subset of checks that best predict the
occurrence of unexpected outcomes. The procedure, based on simulations
of the original data, compares the accuracy and redundancy of different
combinations of analysis validation checks. Accuracy is determined using
binary classification metrics (precision and recall) from a logic
regression model \citep{ruczinski_logic_2003}, while redundancy is
measured using mutual information. The proposed workflow offers a
numerical guarantee that the analysis will produce the expected results,
assuming the assumptions about the data generating mechanism hold.

@@

While some checks may be crucial and directly indicate an unexpected
outcome, others may be tangential to the problem at hand and not
indicate a root cause of an unexpected outcome. In this section, we
propose a procedure to measure the effectiveness of checks that, when
combined using logical operators, contribute to an unexpected outcome. A
small set of independent checks is considered effective if it translates
to unexpected outcomes.

The approach relies on the use of simulated datasets that are generated
based on the analyst's knowledge and assumptions about the data
generation mechanism. Datasets are simulated to have the same structure
and characteristics of the observed data that will be analyzed at some
point in the future. For each simulated dataset, we can apply a
collection of analysis validation checks to the dataset and record which
ones were TRUE and which were FALSE. We can also compute the outcome of
the analysis to see whether the outcome is unexpected in a given
simulated datsets. We can then generate many datasets, each time
applying the analysis validation checks and computing the outcome. After
generating many datasets, we can relate the patterns in the analysis
validation checks to the likely that the outcome will be unexpected in a
given dataset. Figure~\ref{fig-metric-calc} provides an overview of the
process.

@@

From the simulated data, the accuracy branch in
Figure~\ref{fig-metric-calc} refers to a set of checks' ability to
accurately detect unexpected outcomes while minimizing false positives
and false negatives. While a false positive can raise caution or
skepticism about the data, the presence of both false positives and
false negatives suggests that the checks are overly sensitive or lack
sensitivity to unexpected outcomes, respectively.

To incorporate the effect of multiple checks on the outcome, a logic
regression model \citep{ruczinski_logic_2003} is fitted to the analysis
validation checks. Originally developed for SNP microarray data, logic
regression constructs Boolean combinations of binary variables,
\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict both binary and
numerical outcomes. In our case,

\begin{itemize}
\tightlist
\item
  the outcome, \(Y\), is a binary variable indicating whether the result
  is \emph{unexpected} (1) or \emph{expected} (0),
\item
  \(X_1, X_2, \cdots, X_n\) represent the binary analysis validation
  checks, where each check is labeled as 1 if it fails and 0 if it
  passes.
\end{itemize}

For classification problems, the logic regression model is reduced to
the following form:

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\] The logical operations of
the binary variables \(X_1, X_2, \cdots, X_n\), including AND
(\(\land\)), OR (\(\lor\)), and NOT (\(\neg\)) and an example of such
logic combination can be \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\).
Simulated annealing is used to explore the search space by minimizing
the mis-classification error. The following 6 moves are permitted during
the search (as detailed in Figure 2 of \citet{ruczinski_logic_2003}): 1)
replacing a leaf node, 2) replacing an operator, 3) growing a branch, 4)
pruning a branch, 5) splitting a leaf node, and 6) deleting a leaf node.

Compared to other tree-based methods for binary-binary prediction, the
Boolean combinations from the logic regression model produce a tree
structure that can be directly interpreted as the possible combination
of checks leading to an unexpected outcome, without the need to invert
the tree as required in classic tree-based recursive partitioning
methods. The logic regression model is then used to predict the analysis
result based on the values of checks, and the prediction is compared to
the actual analysis result in order to calculate the precision and
recall of the checks.

While checks may score high on predictive accuracy, they may be less
effective at explaining the various reasons behind unexpected results.
This could happen if, for example, a set of checks are all tangentially
related to the cause of the unexpected results, but none addresses the
root cause. It may also occur if the checks are highly correlated with
one another, leading to redundancy.

To quantify redundancy, the concept of mutual information is used.
Mutual information \(I(x, y)\) measures the amount of information shared
between two random variables and is defined as the Kullback-Leibler
distance \(D(p \parallel q)\) between the joint distribution of the two
variables and the product of the marginal distributions:

\[I(x,y) = D\big(p(x,y) \parallel p(x)p(y)\big) = \sum_x \sum_y p(x,y) \log \frac{p(x,y)}{p(x)p(y)}\]

This concept extends naturally to multiple variables through total
correlation, \(C(X_1, X_2, \cdots, X_n)\), which captures redundancy
across a set of \(n\) variables:

@@

A high mutual information value indicates redundancy among the checks,
while a low value suggests that the checks are independent and provides
unique information to diagnose the unexpected outcome. To standardize
this measure, the total correlation \emph{per observation} is
calculated, and an independence score, ranging between 0 and 1, is
defined as 1 - mutual information.

To combine precision, recall, and independence into a single metric, we
can combine the three scores using the arithmetic mean, harmonic mean,
or quadratic mean. The differences among these means are minimal when
the three metrics are similar. However, as the differences among the
metrics increases, the harmonic mean tends to produce the smallest
overall score, as it penalizes low values, while the quadratic mean
tends to produce the largest score by rewarding higher values more. For
simple interpretation of the score, the arithmetic mean is preferred,
while in applications where the difference between precision, recall,
and independence need to be penalized or rewarded more, the harmonic and
quadratic mean should be considered.

@@

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three analysis checks
described previously to generate the prediction of the unexpected
outcome, whether the average number of step falls within the {[}8,500,
9,500{]} interval. The predictions from the logic regression model can
be compared with the simulated true outcome for calculating the
precision and recall metrics. Figure~\ref{fig-logic-reg} shows the
best-fitting logic regression model as

(quantile(step, 0.6) \textgreater{} 10,000 OR quantile(step, 0.4)
\textless{} 8,000) AND sd(step) \textless{} 2,500

In other words, we would predict an unexpected outcome in the analysis
if the standard deviation of the step count is not too large (2,500) and
either the 60\% quantile of the step count exceeds 10,000 or the 40\%
quantile of the step count falls below 8,000.

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the check rule found by
the logic regression. The harmonic and arithmetic means are included to
combine the three measures. The results show that the three checks
produced by the logic regression can accurately predict 86.1\% cases of
all \emph{actual unexpected results} in the simulation data.
Furthermore, 43.1\% of all \emph{predicted unexpected results} were in
fact observed to be unexpected.

@@

\caption{\label{tbl-logic-reg}Accuracy (precision and recall) and
parsimony (independence) metrics for each individual check and the logic
regression check rule. The harmonic and arithmetic means of the three
metrics are included to evaluate the quality of the checks in diagnosing
unexpected step counts (more than five days with fewer than 8,000
steps).}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{17em}rrrrr}
\toprule
checks & precision & recall & indep. & harmonic & arithmetic\\
\midrule
Check 1: q(step, 0.6) > 10000 & 0.319 & 0.575 & 1.000 & 0.511 & 0.631\\
Check 2: q(step, 0.4) < 8000 & 0.264 & 0.613 & 1.000 & 0.467 & 0.626\\
Check 3: sd(step) > 2500 & 0.153 & 0.289 & 1.000 & 0.273 & 0.481\\
Logic regression: (check 1 OR check 2) AND (not check 3) & 0.861 & 0.431 & 0.998 & 0.669 & 0.763\\
Regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774\\
\bottomrule
\end{tabular}}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{18em}rrrrr}
\toprule
check & precision & recall & indep. & harmonic & arithmetic\\
\midrule
Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1 & 0.763 & 0.815\\
Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1 & 0.472 & 0.612\\
Check 3: mortality outlier & 0.302 & 0.775 & 1 & 0.536 & 0.692\\
Check 4: PM10 outlier & 0.305 & 0.766 & 1 & 0.537 & 0.690\\
Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.722 & 0.876 & 1 & 0.851 & 0.866\\
\bottomrule
\end{tabular}}

>>>>

\thanks{CONTACT: H. Sherry
Zhang. Email: \href{mailto:huize.zhang@austin.utexas.edu}{\nolinkurl{huize.zhang@austin.utexas.edu}}. Roger
D.
Peng. Email: \href{mailto:roger.peng@austin.utexas.edu}{\nolinkurl{roger.peng@austin.utexas.edu}}. }
\begin{document}
\captionsetup{labelsep=space}
\maketitle
\textsuperscript{1} Department of Statistics and Data
Sciences, University of Texas at Austin, Texas, United States
\begin{abstract}
In data analysis, unexpected results often prompt researchers to revisit
their procedures to identify potential issues. While some researchers
may struggle to identify the root causes, experienced researchers can
often quickly diagnose problems by checking a few key assumptions. These
checked assumptions, or expectations, are typically informal, difficult
to trace, and rarely discussed in publications. In this paper, we
introduce the term \emph{analysis validation checks} to formalize and
externalize these informal assumptions. We then introduce a procedure to
identify a subset of checks that best predict the occurrence of
unexpected outcomes, based on simulations of the original data. The
checks are evaluated in terms of accuracy, determined by binary
classification metrics, and independence, which measures the shared
information among checks. We demonstrate this approach with a toy
example using step count data and a generalized linear model example
examining the effect of particulate matter air pollution on daily
mortality.
\end{abstract}

@@

In this paper, we conceptualize these inexplicit expectations as
\emph{analysis validation checks}, which allows us to examine the
assumptions made about the data during an analysis. We then introduce a
procedure to identify the subset of checks that best predict the
occurrence of unexpected outcomes. The procedure, based on simulations
of the original data, compares the accuracy and independence of
different combinations of analysis validation checks. Accuracy is
determined using binary classification metrics (precision and recall)
from a logic regression model \citep{ruczinski_logic_2003}, while
independence is measures the shared information among checks. The
proposed workflow offers a numerical guarantee that the analysis will
produce the expected results, assuming the assumptions about the data
generating mechanism hold.

@@

While not all analysis validation checks are equally important, some may
be more useful at diagnosing unexpected outcomes than others. This
section introduces a procedure to identify a subset of analysis
validation checks that best predict the occurence of unexpected
outcomes.

The approach uses simulated datasets that have the same structure and
characteristics of the observed data to be analyzed in the future. For
each simulated dataset, analysis validation checks are applied to record
their TRUE or FALSE values. The outcome of the analysis is also computed
to determine whether it is unexpected (1) or expected (0). By generating
many datasets, each time applying the analysis validation checks and
computing the outcome, we can relate the patterns in the checks to the
likely that the outcome will be unexpected. Figure~\ref{fig-metric-calc}
provides an overview of the process.

@@

\subsection{Accuracy}\label{accuracy}

From the simulated data, the accuracy branch in
Figure~\ref{fig-metric-calc} refers to the set of checks' ability to
detect unexpected outcomes accurately. To relate multiple checks to the
outcome, a logic regression model \citep{ruczinski_logic_2003} is used.
Originally developed for SNP microarray data, logic regression
constructs Boolean combinations of binary variables,
\(\mathcal{L}(X_1, X_2, \cdots, X_n)\), to predict classification or
regression outcomes. The logical operations allowed to construct
\(\mathcal{L(\cdot)}\) includes AND (\(\land\)), OR (\(\lor\)), and NOT
(\(\neg\)) and an example of the Boolean combination of
\(\mathcal{L(\cdot)}\) is \(X_1 \text{ AND } (X_2 \text{ OR } X_3)\). In
our binary-binary classification problem, the logic regression model can
be written as follows:

\[E(Y) = \mathcal{L}(X_1, X_2, \cdots, X_n)\]

where the outcome, \(Y\), indicates whether the outcome is
\emph{unexpected} (1) or \emph{expected} (0), and the analysis
validation checks, \(X_1, X_2, \cdots, X_n\), are labeled as 1 if they
fail and 0 if they pass.

This objective function is optimized using simulated annealing algorithm
to minimize the misclassification error and the following 6 moves are
permitted during the optimization to grow or purne the tree (as detailed
in Figure 2 of \citet{ruczinski_logic_2003}): 1) replacing a leaf node,
2) replacing an operator, 3) growing a branch, 4) pruning a branch, 5)
splitting a leaf node, and 6) deleting a leaf node. Predictions from the
logic regression model is then used, along with the observed outcome, to
calculate the precision and recall of the optimized Boolean combination
of checks. Compared to other tree-based methods for binary-binary
prediction, the Boolean combinations from the logic regression model
produce a tree structure that can be directly interpreted as the
possible combination of checks leading to an unexpected outcome, without
the need to invert the tree as required in classic tree-based recursive
partitioning methods.

\subsection{Independence}\label{independence}

While checks may score high on predictive accuracy, they may be
correlated and less useful to provide actions to act on the analysis.
This could happen if a set of checks are all tangentially related to the
cause of the unexpected results, but none addresses the root cause. The
independence score in Figure~\ref{fig-metric-calc} is defined as 1 -
total correlation per observation.

Total correlation is a generalization of mutual information that
measures the amount of shared information across multiple variables
multiple variables based on Kullback-Leibler distance:

@@

A high total correlation value indicates overlapping of information
among the analysis validation checks, while a low value suggests that
the checks are independent and provides unique information to diagnose
the unexpected outcome. To standardize this measure, the total
correlation \emph{per observation} is used in calculating the
independence score.

The three metrics (precision, recall, and independence) can be combiend
into a single metric using the arithmetic mean, harmonic mean, or
quadratic mean. The differences among these means are minimal when the
three metrics are similar. However, as the differences among the metrics
increases, the harmonic mean tends to produce the smallest overall
score, as it penalizes low values, while the quadratic mean tends to
produce the largest score by rewarding higher values more. For simple
interpretation of the score, the arithmetic mean is preferred, while in
applications where the difference between precision, recall, and
independence need to be penalized or rewarded more, the harmonic and
quadratic mean should be considered.

@@

Returning to the step count example introduced in Section~\ref{sec-toy},
the logic regression model is fitted to the three validation checks
described previously to predict the outcome, whether the average number
of step falls within the {[}8,500, 9,500{]} interval.
Figure~\ref{fig-logic-reg} shows the best-fitting logic regression model
as

\begin{quote}
(quantile(step, 0.6) \textgreater{} 10,000 OR quantile(step, 0.4)
\textless{} 8,000) AND (NOT sd(step) \textgreater{} 2,500)
\end{quote}

We would predict an unexpected outcome in the analysis if the standard
deviation of the step count is not too large (2,500) and either the 60\%
quantile of the step count exceeds 10,000 or the 40\% quantile of the
step count falls below 8,000.

@@

Table~\ref{tbl-logic-reg} presents the calculated precision, recall, and
independence for the three individual checks and the check rule found by
the logic regression. The harmonic and arithmetic means are included to
combine the three measures. The results show that the three checks
produced by the logic regression can accurately predict 86.1\% cases of
all \emph{predicted unexpected results} in the simulation data.
Furthermore, 43.1\% of all \emph{actual unexpected results} were in fact
observed to be unexpected.

@@

\caption{\label{tbl-logic-reg}Precision, recall, and independence
calculated for each individual check and the logic regression check
rule. The harmonic and arithmetic means of the three metrics are
included to evaluate the quality of the checks in diagnosing unexpected
step counts (more than five days with fewer than 8,000 steps).}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{17em}rrrrr}
\toprule
Checks & Precision & Recall & Independence & Harmonic & Arithmetic\\
\midrule
Check 1: q(step, 0.6) > 10000 & 0.319 & 0.575 & 1.000 & 0.511 & 0.631\\
Check 2: q(step, 0.4) < 8000 & 0.264 & 0.613 & 1.000 & 0.467 & 0.626\\
Check 3: sd(step) > 2500 & 0.153 & 0.289 & 1.000 & 0.273 & 0.481\\
Logic regression: (check 1 OR check 2) AND (not check 3) & 0.861 & 0.431 & 0.998 & 0.669 & 0.763\\
Regression tree & 0.542 & 0.780 & 1.000 & 0.727 & 0.774\\
\bottomrule
\end{tabular}}

@@

\centering
\resizebox{\ifdim\width>\linewidth\linewidth\else\width\fi}{!}{
\begin{tabular}{>{\raggedright\arraybackslash}p{18em}rrrrr}
\toprule
Checks & Precision & Recall & Independence & Harmonic & Arithmetic\\
\midrule
Check 1: cor(m, PM10) < -0.03 & 0.551 & 0.894 & 1 & 0.763 & 0.815\\
Check 2: cor(m, tmp) > -0.35 & 0.281 & 0.556 & 1 & 0.472 & 0.612\\
Check 3: mortality outlier & 0.302 & 0.775 & 1 & 0.536 & 0.692\\
Check 4: PM10 outlier & 0.305 & 0.766 & 1 & 0.537 & 0.690\\
Logic regression: (check 1) AND ((check 2) OR (check 3 AND check 4)) & 0.722 & 0.876 & 1 & 0.851 & 0.866\\
\bottomrule
\end{tabular}}
